---
title: "FAA Dataset Analysis"
subtitle: "Analysis of Aircraft Overruns"
author: "Meenal Narsinghani"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 6
    toc_float: true
    collapsed: false
---

***

### Initial Setup


#### Loading the libraries
```{r, message=FALSE, error=FALSE, warning=FALSE}
library(gdata)
library(readxl)
library(tidyverse)
library(dplyr)
library(MASS)
library(gridExtra)
```


#### Importing the dataset
```{r}
faa1 <- read_excel("FAA1.xls")
head(faa1)

faa2 <- read_excel("FAA2.xls")
head(faa2)
```

***

### Analytical dataset creation

#### Checking the structure of the datasets
```{r}
## For data set FAA1
str(faa1)   ## Sample size: 800; Number of variables = 8

## For data set FAA2
str(faa2)   ## Sample size: 150; Number of variables = 7

## Dataset FAA2 does not have "Duration" variable. 

```


#### Merging the datasets
```{r}
#Since FAA2 has one less variable, creating a new column "Duration" in FAA2 and setting it to NA

faa2$duration <- NA
merged_faa <- rbind(faa1, faa2)
head(merged_faa)

merged_faa$aircraft <- as.factor(merged_faa$aircraft)
str(merged_faa)   ##Total number of observations = 950; Number of variables = 8
```


#### Duplicate records treatment
```{r}
##Checking if duplicates exist
nrow(merged_faa[duplicated(merged_faa[,c(1,3:8)]),])    ## 100 duplicate observations found

##Removing duplicated rows
merged_faa_distinct <- merged_faa[!duplicated(merged_faa[,c(1,3:8)]),]
nrow(merged_faa_distinct)

```


#### Summary statistics
```{r}
str(merged_faa_distinct)
#Number of observations = 850
#Number of variables = 7

#Summary statistics for each variable
summary(merged_faa_distinct)

```


    **Important observations**
    
    1. NAs in speed air
    2. Negative height
    3. unsually large range for distance variable : >6000
    4. Speed_ground < 30 : abnormal
    5. Max speed_ground, speed_air > 140
    6. Duration < 40min
  

#### Data Cleaning and further exploration

**Checking for abnormal values in the data set. We will remove the rows that contain any 'abnormal values'**

    1. Negative value for "Height" variable or where "Height" < 6
    2. Observations with speed_air = NA
    3. Observations with speed_air or speed_ground > 140
    4. Observation with speed_air or speed_ground < 30
    5. Distance > 6000ft


```{r}
faa_data_cleaned <- merged_faa_distinct %>% 
                filter(duration > 40 | is.na(duration)) %>%
                  filter((speed_ground >= 30 && speed_ground <= 140) | is.na(speed_ground))%>%
                    filter((speed_air >= 30 && speed_air <= 140) | is.na(speed_air)) %>% 
                      filter(height >= 6 | is.na(height)) %>% 
                        filter(distance < 6000 | is.na(distance))

nrow(faa_data_cleaned)   #833 obs
```

Total rows removed = 17

*Retaining observations with NAs as dataset FAA2 has no "Duration" column. Dropping all the observations where any missing value is present will render dataset FAA2 as useless.*

***


### Exploratory Data Analysis
```{r}
hist(faa_data_cleaned$no_pasg, main = "Histogram of No. of passengers", xlab = "")
hist(faa_data_cleaned$speed_ground, main = "Histogram of Speed_Ground", xlab = "")
hist(faa_data_cleaned$speed_air, main = "Histogram of Speed_Air", xlab = "" )
hist(faa_data_cleaned$duration,main = "Histogram of Duration", xlab = "")
hist(faa_data_cleaned$height, main = "Histogram of Height", xlab = "")
hist(faa_data_cleaned$pitch, main = "Histogram of Pitch", xlab = "")
hist(faa_data_cleaned$distance, main = "Histogram of Distance", xlab = "")

```

    ** A quick summary of what we have observed so far from our analysis:**
    
    1. 17 observations corresponding to abnormal data values were omitted from the cleaned data set
    2. Columns 'Duration' and 'Speed_Air' has some missing values
    3. From the histograms we observe that distributions of all variables is symmetric(normal), except for "Speed_Air" and "Distance"
    4. There is no observation in the data set where the value of "Speed_Air" is below 90 (allowed min = 30)
    5. Distribution of "Speed_Air" and "Distance" is left skewed

***  

### Factor Analysis

**Initial analysis for identifying important factors that impact the response variable 'landing distance'**

Computing the pairwise correlation between the landing distance and each factor X. 

```{r}
cor_with_dist <- round(cor(faa_data_cleaned$distance, faa_data_cleaned[,-1],use = "pairwise.complete.obs"),2)

cor_with_dist

```


Plotting scatter plots to represent the relationship between the response variable - `distance` and different predictor variables

```{r}

scatter1 <- ggplot(faa_data_cleaned, aes( x = duration, y = distance)) + geom_point()
scatter2 <- ggplot(faa_data_cleaned, aes( x = no_pasg, y = distance)) + geom_point()
scatter3 <- ggplot(faa_data_cleaned, aes( x = speed_ground, y = distance)) + geom_point()
scatter4 <- ggplot(faa_data_cleaned, aes( x = speed_air, y = distance)) + geom_point()
scatter5 <- ggplot(faa_data_cleaned, aes( x = height, y = distance)) + geom_point()
scatter6 <- ggplot(faa_data_cleaned, aes( x = pitch, y = distance)) + geom_point()

grid.arrange(scatter1,scatter2,scatter3, scatter4, scatter5, scatter6, nrow = 2, ncol = 3)

```


Re-coding the `aircraft` from character variable to factor

```{r}
#Code "Boeing" as 1 and "Airbus" as 0
levels(faa_data_cleaned$aircraft) <- c(1,2)

faa_data_cleaned$aircraft <- as.numeric(levels(faa_data_cleaned$aircraft))[faa_data_cleaned$aircraft]

summary(faa_data_cleaned)

cor_with_dist <- round(cor(faa_data_cleaned$distance, faa_data_cleaned[,1:7],use = "pairwise.complete.obs"),2)

cor_with_dist

scatter1 <- ggplot(faa_data_cleaned, aes( x = duration, y = distance)) + geom_point()
scatter2 <- ggplot(faa_data_cleaned, aes( x = no_pasg, y = distance)) + geom_point()
scatter3 <- ggplot(faa_data_cleaned, aes( x = speed_ground, y = distance)) + geom_point()
scatter4 <- ggplot(faa_data_cleaned, aes( x = speed_air, y = distance)) + geom_point()
scatter5 <- ggplot(faa_data_cleaned, aes( x = height, y = distance)) + geom_point()
scatter6 <- ggplot(faa_data_cleaned, aes( x = pitch, y = distance)) + geom_point()
scatter7 <- ggplot(faa_data_cleaned, aes( x = aircraft, y = distance)) + geom_point()

grid.arrange(scatter1,scatter2,scatter3, scatter4, scatter5, scatter6, scatter7,nrow = 3, ncol = 3)


```


***

### Simple Regression - single factor each time

```{r}

linear_model1 <- lm(distance ~ aircraft, data = faa_data_cleaned)
#summary(linear_model1)  #p-value = 4.38e-12

linear_model2 <- lm(distance ~ duration, data = faa_data_cleaned)
#summary(linear_model2)  #p-value = 0.146

linear_model3 <- lm(distance ~ no_pasg, data = faa_data_cleaned)
#summary(linear_model3)  #p-value = 0.618

linear_model4 <- lm(distance ~ speed_ground, data = faa_data_cleaned)
#summary(linear_model4)  #p-value = <2e-16

linear_model5 <- lm(distance ~ speed_air, data = faa_data_cleaned)
#summary(linear_model5)  #p-value = <2e-16

linear_model6 <- lm(distance ~ height, data = faa_data_cleaned)
#summary(linear_model6)  #p-value = 0.00389

linear_model7 <- lm(distance ~ pitch, data = faa_data_cleaned)
#summary(linear_model7)  #p-value = 0.0127


#With all the variables
linear_model <- lm(distance ~ ., data = faa_data_cleaned)
summary(linear_model)   ##Aircraft, speed_air, height

```

***

### Simple Regression - using standardized variables

```{r}

stnd_var <- function(var)
{
  var <- (var-mean(var,na.rm = T))/sd(var,na.rm = T)
}

faa_data_stnd <- sapply(faa_data_cleaned,stnd_var)
#head(faa_data_stnd)

faa_data_stnd <- as.data.frame(faa_data_stnd, row.names = NULL)

cor_with_dist_stnd <- round(cor(faa_data_stnd$distance, faa_data_stnd[,1:7],use = "pairwise.complete.obs"),2)


cor_with_dist_stnd

```


**From the correlation values and the results of the simple regression analysis we see that following are the significant variables - **

    1. speed_air
    2. speed_ground
    3. aircraft
    4. height
    5. duration
    6. pitch
    7. no_pasg


***

#### Checking for collinearity

We will compare the regression coefficients of the three models below:
Model 1: LD ~ Speed_ground
Model 2: LD ~ Speed_air
Model 3: LD ~ Speed_ground + Speed_air

to check for the correlation between Speed_ground and Speed_air. 

```{r}
model1 <- lm(distance~speed_ground, data = faa_data_cleaned, na.action = na.omit)
summary(model1)

model2 <- lm(distance~speed_air, data = faa_data_cleaned)
summary(model2)

model3 <- lm(distance~speed_ground + speed_air, data = faa_data_cleaned)
summary(model3)
```


We see that speed_ground has changed sign to negative. Also, it has become insignificant. 

**Since the correlation between speed_ground and speed_air is very high. We would want to retain speed_ground**

***

**Based on the results we observe following variables are significant in predicting landing distance:**
**(variables arranged in decreasing order of significance)**

    1. speed_ground
    2. aircraft
    3. height
    4. duration
    5. pitch
    6. no_pasg


We will fit the following six models:
Model 1: LD ~ X1
Model 2: LD ~ X1 + X2
Model 3: LD ~ X1 + X2 + X3
....

where X1, X2... are predictors arranged in decreaasing order of significance

```{r, message=FALSE, error=FALSE, warning=FALSE}
rsquared <- NULL

model1 <- lm(distance~speed_ground, data = faa_data_cleaned)
rsquared[1] <- summary(model1)$r.squared

model2 <- lm(distance~speed_ground + aircraft, data = faa_data_cleaned)
rsquared[2] <- summary(model2)$r.squared

model3 <- lm(distance~speed_ground + aircraft + height, data = faa_data_cleaned)
rsquared[3] <- summary(model3)$r.squared

model4 <- lm(distance~speed_ground + aircraft + height + duration, data = faa_data_cleaned)
rsquared[4] <- summary(model4)$r.squared

model5 <- lm(distance~speed_ground + aircraft + height + duration + pitch, data = faa_data_cleaned)
rsquared[5] <- summary(model5)$r.squared

model6 <- lm(distance~speed_ground + aircraft + height + duration + pitch + no_pasg, data = faa_data_cleaned)
rsquared[6] <- summary(model6)$r.squared
```

**We will not plot the variation of R-squared against number of predictors(p)**
```{r}
plot(1:6, rsquared,type = 'b', main = 'Variation of R-sqaured against number of predictors (p)')
```


***

**Instead of R-squared, lets determine the `adjusted R-sqaured` value and observe its pattern against number of predictors**
```{r, message=FALSE, error=FALSE, warning=FALSE}
adj_rsquared <- NULL

adj_rsquared[1] <- summary(model1)$adj.r.squared
adj_rsquared[2] <- summary(model2)$adj.r.squared
adj_rsquared[3] <- summary(model3)$adj.r.squared
adj_rsquared[4] <- summary(model4)$adj.r.squared
adj_rsquared[5] <- summary(model5)$adj.r.squared
adj_rsquared[6] <- summary(model6)$adj.r.squared
```


```{r}
plot(1:6, adj_rsquared, type = "b", main = 'Variation of adjusted R-sqaured against number of predictors (p)')
```


***

**Determining the pattern of AIC against number of predictors p**
```{r, message=FALSE, error=FALSE, warning=FALSE}
aic <- NULL

aic[1] <- AIC(model1)
aic[2] <- AIC(model2)
aic[3] <- AIC(model3)
aic[4] <- AIC(model4)
aic[5] <- AIC(model5)
aic[6] <- AIC(model6)
```


```{r}
plot(1:6, aic, type = 'b', main = 'Variation of AIC against number of predictors (p)')
```

**From the above 3 plots, it is appropriate to select speed_ground, aircraft make and height to build the final model.**

*Adding more variables will increase the complexity of the model and will provide any additional information to the model*

***

### Variable selection - Forward StepAIC

```{r}
full_model <- lm(distance ~ .-speed_air , data = faa_data_cleaned)

step.model <- stepAIC(full_model, direction = "forward", 
                      trace = FALSE)
summary(step.model)

```

```{r}
#Final model

final_model <- lm(distance ~ aircraft + speed_ground + height, data = faa_data_cleaned)
summary(final_model)
```

